I would like to thank my parents, Lizzie and Andrew, who brought me up in a household where curiosity was encouraged. Dad, your bedtime stories about the Big Bang, entangled quarks (I imagined them as ducks flipping in unison) and black holes put me on this path. And Mum, I could not have come this far without your infectious enthusiasm, your drive to organise specialised tuition, your motivation to take part in science competitions and events, and importantly, your encouragement for me to have other hobbies! Thank you both for supporting me intellectually, emotionally (and financially!) throughout my education. I hope that despite your differences you can be united in sharing the pride of this achievement with me. Thanks also to my siblings, Olly and Margaux, whose friendship and humour have made the journey so enjoyable.

Thank you to my friends: those outside of academia, who have had to endure my endless ramblings about the glory of science; and those inside of academia, who have had to endure my endless ramblings about the dire state of politics. In particular, cheers to Adam, Adinda, Federico and Matthew for keeping me sane during our LTA. Although I'm still upset I never managed to win a game of Catan against you.

Thank you Paul and Chris, the best supervisors I could have hoped for: you provided fascinating projects to work on; always had time to give insightful feedback; and were \emph{very} generous with beer and coffee. On that topic, cheers to Seth, who was equally generous. Thanks for teaching me how to be a researcher, and for making Physics fun through the good times (\&) and the bad (holidayspirit). I'd like to express my gratitude to the IC HEP group and the CMS collaboration for providing an outstanding environment to do science in, and to the STFC for funding my studies and LTA. 

Finally, thank you to my fianc\'ee Emma, whose love and support have guided me through the highs and lows. Thank you for enduring the stress, the intervals of spatial separation and the ruined holidays. Conversely, thank you for enjoying the adventure and excitement, the new experiences and the skiing. Above all, thank you for being proud of my research and achievements, and for pushing me to follow my dreams. 
